\addcontentsline{toc}{section}{Introduction}
\section*{Introduction}

Free groups were first studied in \cite{Fuchs1980} as an example of Fuchsian groups, which are discrete subgroups of $PSL_2(\Z)$, but they did not become relevant on their own until the late 1800s. Any group may be written using its group presentation (for example, $D_3 = \langle r,s | r^3 = s^2 = 1, rs = sr^2 \rangle$). In \cite{Dyck1882}, von Dyck, who was first to express groups using a presentation, showed that free groups have the simplest possible one ($n$ generators with no relations) and that all finitely generated groups are the quotient of a free group. The algebraic properties of free groups were investigated by Nielsen, who gave them their name in \cite{Nielsen1924}, while Dehn, Reidemeister, and Schreier also played important roles in developing the field throughout the first half of the 1900s. 

The simple nature of free groups is equal parts useful and deceiving. Consider The Word Problem. 

\vspace{0.5cm}

\begin{center}
    \begin{minipage}{0.7\textwidth}
        \textbf{The Word Problem}: For a group $G$, let $g$ be an element expressed as a product of $G$'s generators. Find an algorithm that shows, in finitely many steps, whether or not $g$ is the identity element.
    \end{minipage}
\end{center}

\vspace{0.5cm}

While it is known that this problem is undecidable for arbitary groups, we will show in the next section that it is solvable if $G$ is a free group. This does not mean that free groups are a trivial class of groups. In fact, understanding their algebraic properties has been an area of study for over a hundred years. To demonstrate this, we consider two questions.

First, suppose we are given the presentation of an arbitrary group $G$. Can we easily check whether or not $G$ is a free group? This question was answered in the negative in \cite{Adyan1955,Rabin1958}, where it was shown that such a question is undecidable. 

The second question is one that is often asked when a property of a structure is first defined: do the substructures of this structure also have this property? In particular, are the subgroups of a free group also free groups? This was answered in the positive in the 1920s, first in \cite{Nielsen1921} for finitely generated subgroups, then in \cite{Schreier1928} in the full form. However, such proofs were not trivial, and required high level topological or algebraic methods to get to a rigorous proof. Understanding the proofs would require understanding very high level results in group theory, far beyond the level of an undegraduate course in abstract algebra. 

Answers to the both of these questions do exist in simpler forms within the field of Combinatorial/Geometric Group Theory, which in part seeks to understand a group's algebraic properties by looking at how it acts on a space. This extra assumption on how a group acts upon a space makes these results much easier to understand at the undergraduate level. 

An answer to the first question can be found thanks to the Ping-Pong Lemma, which describes a sufficient condition for a group to be free, so long as it acts in a certain way on a set $X$. The result is often attributed to Felix Klein \cite{delaHarpe2000}, who used it in the late 1800s, but is also called Schottky's Lemma, after Freidrich Schottky. 

\begin{thm}[The Ping-Pong Lemma]
    Let $\{g_1, g_2, \ldots, g_n\}$ generate a group $G$, which acts on a set $X$. If 

    \begin{enumerate}[label=(\roman*)]
        \item $X$ contains $n$ subsets $X_1, \ldots, X_n$ such that $X_i \cap X_j = \varnothing$, and
        \item $g_i^k(X_j) \subset X_i$ for all nonzero powers $k$ and $i \neq j$.
    \end{enumerate}

    Then $G$ is isomorphic to $F_n$. 
\end{thm}

More interestingly, an answer to question two is a corollary of a result proven by Serre in \cite{Serre1980}, who showed that free groups may be classified by how they act on trees.

\begin{thm}[Serre's Theorem, 1980]
    $G$ is a free group if and only if it acts freely on a tree without inversion. 
\end{thm}

In this paper, we present full proofs and explanations of both results. First, we prove the Ping-Pong Lemma and use it to show that free groups of any rank, including one of infinite rank, lie within the matrix group $SL_2(\Z)$. Afterwards, we prove both sides of Serre's Theorem. In proving the forwards direction, we define a special graph for our free group to act on called the Cayley graph, and then show that free groups acts freely on their corresponding Cayley graphs. To prove the reverse direction, we show that every group acting on a tree induces a special tiling of the tree, before finding a free generating set using this tiling. 