\chapter{Week 1}

Throughout this course, we will explore several different concepts that will, when combined, make up our theory of algebraic curves. Some of the concepts we will talk about will include:

\begin{itemize}
    \item Projective Geometry
    \item B\'{e}zout's Theorem about intersections of curves in the plane.
    \item Riemann Surfaces, which are related to algebraic curves over $\mathbb{C}$. These will help motivate some ideas we will see much later on, such as differentials.
    \item Curves over finite fields, including elliptic curves, which are related to concepts in number theory.
\end{itemize}

\section{A First Glimpse at Projective Geometry}

\subsection{The Projective Plane in 2 Dimensions}

Let $\mathbb{F}$ be any field. Then $\mathbb{F}^3$ is a 3D space. Notice that the 2D space $\mathbb{F}^2$ is contained within $\mathbb{F}^3$ as a slice of the 3D space. An explicit map may be constructed as

\[(x,y) \mapsto (x,y,1)\]

There is thus a projection of the 2D space into the 3D space. 

Now, consider the set of all lines in $\mathbb{F}^3$ that pass through the origin. They all meet at the origin, and nowhere else. This means any point in $\mathbb{F}^3 \setminus \{0\}$ corresponds to a unique line in this space. If we scale this point, we stay on the line. Thus, if we quotient by the set of scalars, also known as $\mathbb{F}^*$, then we will get a set of points that correspond uniquely to a single line in the 3D space. This set,

\[(\mathbb{F}^3 \setminus \{0\})/\mathbb{F}^*\]

is known as the \textbf{Projective Plane}, and is denoted by $\mathbb{P}^2$. Going back to our $\mathbb{F}^2$ discussion, any $(x,y) \in \mathbb{F}^2$ can be thought of as the point $(x,y,1) \in \mathbb{F}^3$. Taking the representative $[(x,y,1)] \in \mathbb{P}^2$, we get that each point in $\mathbb{F}^2$ corresponds to a point in $\mathbb{P}^2$, and so $\mathbb{F}^2 \subset \mathbb{P}^2$. The set $\mathbb{P}^2 \setminus \mathbb{F}^2$ only contains the points of the form $[(x,y,0)]$.

To visualize the projective plane, consider the below figure:

The square $\mathbb{F}^2$ is the \textbf{affine plane}, whilst the point $\mathbb{F}^0$ and the line $\mathbb{F}^1$ combine to become the \textbf{line at infinity}.

Let's use a finite field $\mathbb{F}_q$. The set $\mathbb{F}_q^3 \setminus\{0\}$ has $q^3 - 1$ points, while $\mathbb{F}_q^*$ has $q-1$ points. Thus, the projective plane with repsect to $\mathbb{F}_q$ has

\[\frac{q^3 - 1}{q-1} = q^2 + q + 1\]

points. In particular, there are $q^2$ points of the form $(a,b,1)$, $q$ points of the form $(a,1,0)$, and 1 point of the form $(1,0,0)$. 

The exponent should make clear that $\mathbb{P}^2$ is a 2D space. There is a nice correlation between the geometry in the projective plane and that of the underlying field. By construction, any point in $\mathbb{P}^2$ is a line in $\mathbb{F}^3$ passing through the origin. Similarly, any line in $\mathbb{P}^2$ is a plane in $\mathbb{F}^3$ that passes through the origin. For example, consider the plane $ax + by + cz = 0$ in $\mathbb{F}^3$. This corresponds to the line $(a,b,c)$ in $\mathbb{P}^2$. The set of all lines in $\mathbb{P}^2$ is given by the set

\[(\mathbb{F}^3) \setminus \{0\}/\text{scaling}\]

This idea that connects lines and points is known as \textbf{line-point duality}, and is very important. 

Another way to see the connection between lines and points is through a graph. Consider two sets of vertices, one for points in $\mathbb{P}^2$, and one for lines in the same space, where the underlying field is $\mathbb{F}_q$. An edge connects any point to any line that the point lies on. There are $q^2 + q + 1$ vertices in each set, and each vertex has degree $q-1$. This graph is bipartite, and is the largest bipartite graph with this number edges that does not contain a $K_{2,2}$ subgraph. This is because two lines cannot intersect at two distinct points.

\subsection{The Projective Plane in n Dimensions}

We can generalize what we just did to $n$ dimensions to get $\mathbb{P}^n$, which is defined to be

\[\mathbb{P}^n = (\mathbb{F}^{n+1} \setminus \{0\})/\mathbb{F}^*\]

Over a finite field $\mathbb{F}_q$, we get that

\[|\mathbb{P}^n| = q^n + q^{n-1} + \cdots + 1\]

where each $q^k$ is a copy of $\mathbb{F}_q^k$. Instead of lines, we are now considering \textbf{hyperplanes}, which are codimension 1 subspaces of $\mathbb{F}^{n+1}$, that pass through the origin. Similar to before, a point in $\mathbb{P}^n$ corresponds to a hyperplane in $\mathbb{F}^{n+1}$, given by an equation 

\[a_0x_0 + \cdots a_nx_n = 0\]

The space of all hyperplanes is also a $\mathbb{P}^n$. This gives us \textbf{hyperplane-point duality}.

\subsection{The Riemann Sphere}

There are some operations that allow us to take equations in the affine plane and convert them into equations in the projective space. Let's consider $\mathbb{P}^1$, where the underlying field is $\mathbb{C}$. This is the set of pairs $(a,b) \in \mathbb{C}^2 \setminus \{0\}$, modulo scaling. One can think of this space as the set

\[\mathbb{P}^1 = (\mathbb{C} \times \{1\}) \cup (1,0) \cong \mathbb{C} \cup \{\infty\}\]

This can be thought of as the complex numbers, with a single point, the point at infinity, lying above it. If we fold $\mathbb{C}$ onto this point at infinity, we get a sphere, called the \textbf{Riemann Sphere}: 

To go from one to the other, we use \textbf{stereographic projection}. If we define the point at inifnity as the north pole $N$, we get a map which sends $S^2\setminus N \to \mathbb{C}$, and $N \to (1,0)$.

\section{Algebraic Sets}

We now begin introducing some relevant concepts in algebraic geometry. We let $\mathbb{F}$ be an algebraically closed field, and consider $\mathbb{F}[x_1,\ldots,x_n]$, the field of polynomials in $n$ variables with coefficients in $\mathbb{F}$. 

We say that $S \subseteq \mathbb{F}^n$ is an \textbf{algebraic set} if there exists a set $U \subseteq \mathbb{F}[x_1,\ldots,x_n]$ such that we can write 

\[S = \{x \in \mathbb{F}^n : f(x) = 0 \quad \forall f \in U\}\]

meaning $S$ vanishes for every polynomial in $U$. 

We should note some important facts:

\begin{enumerate}
    \item The algebraic sets in $\mathbb{C}$ are $\mathbb{C}$, and all finite subsets.
    \item Every finite subset of $\mathbb{C}^n$ is algebraic: For a point $(\alpha_1, \ldots, \alpha_n)$, take $U = \{x_1 - \alpha_1, \ldots, x_n - \alpha_n\}$. 
\end{enumerate}

Now, we want to pick $U$ so that it is as large as possible. For an algebraic set $S$, the \textbf{ideal} of $S$ is the set $I(S)$ of all polynomials that are equal to 0 on $S$. Recall by the Hilbert Basis Theorem that $\mathbb{F}[x_1,\ldots, x_n]$ is Noetherian, and thus $I(S)$ is always finitely generated. This means that there are polynomials $p_1, \ldots, p_k$ for which

\[I(S) = \{\sum_i A_i p_i : A_i \in \mathbb{F}[x_1,\ldots, x_n]\}\]

We can also go in the other direction. For any ideal $I \subseteq \mathbb{F}[x_1,\ldots, x_n]$, we can define the \textbf{vanishing set} $V(I)$ to be the set

\[\{x \in \mathbb{F}^n : P(x) = 0 \quad \forall P \in I\}\]

which is all values that map to 0 for every polynomial in $I$. This gives a nice duality: If $S$ is algebraic, then $V(I(S)) = S$, and for an ideal $J$, 

\[I(V(J)) = \sqrt{J} = \{P(x_1,\ldots,x_n) : P^m \in J \, \text{ for some $m$}\}\]

This idea that $I(V(J)) = \sqrt{J}$ is refered to as \textbf{Hilbert's Nullstellensatz}, or Hilbert's theorem of zeroes. It allows us to translate a problem about algebraic curves or surfaces into a question about algebra. 

Given an algebraic set $S$, we can also form the ring 

\[\mathbb{F}[x_1,\ldots,x_n]/I(S)\]

called the \textbf{coordinate ring of $S$}.

Let's consider an example to see everything in action. In $\mathbb{C}^2$, consider the curve 

\[y^2 = x^3 + 1\]

which is the polynomial $y^2 - x^3 + 1$. We take $S = V(y^2 - x^3 - 1)$. The set 

\[\{Q(x,y)\big|_S\} = \{f(x) + yg(x) : f,g \in \mathbb{C}[x]\}\]

is the coordinate ring of $S$ with operations modulo $y^2 - x^3 + 1$, since all polynomials will be linear in $y$. 

Notice that, in a small enough neighbourhood of $x = 0$, we can write $y = \sqrt{x^3 + 1}$, which equals the power series

\[1 + \frac{x^3}{2} + \cdots \]

or the negative of that power series. This corresponds to the fact that there are two points at which the line $x = 0$ and our curve intersect. To see this more explictely, we can notice that 

\begin{align*}
    \mathbb{C}[x,y]/(y^2 - x^3 - 1, x) & = \mathbb{C}[x,y]/(y^2 - 1) \\
    & = \mathbb{C}[x,y]/(y-1) \oplus \mathbb{C}[x,y]/(y+1) \\
    & \cong \mathbb{C}^2
\end{align*}

as it is 2D, we get 2 solutions. 

Note that this is not always something we can do. For example, near $x = 0$, the curve $y^2 = x^3 + x^2$ looks like the curve $y^2 = x^2$, which is a cross, giving us a singularity. 